Supplementary Information

### 2021-02-18 writing/transect_paper.Rmd
Additional changes from Martin

Sampling was conducted at two sites, near the villages of Angaradebou (Ang: N 9.75456° E 2.14064°) and Gando (Gan: N 9.75678° E 2.31058°) approximately 30\ km apart in the _Forêt Classée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in West Sudanian savannah woodlands [@olson2001;@yorou2014] dominated by the ECM host tree _Isoberlinia doka_ (Fabaceae tr. Amherstieae).
At each site, 25\ soil samples were collected along a single 24\ m linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016, for a total of 67 samples.
For each sample, coarse organic debris was removed from the soil surface and a sample of approximately 5\ cm\ × 5\ cm\ × 5\ cm was extracted with an ethanol sterilized knife blade and homogenized by manually breaking up clumps and mixing inside a plastic zipper bag.
A subsample of approximately 250\ mg was preserved and returned to the laboratory for extraction (see Supplementary Methods \@ref(DNA-supplement) for preservation and extraction methods).

@@

We conclude that this poor performance was due to a combination of long amplicon length and low sequencing depth relative to community diversity (see Supplementary Methods \@ref(long-numbers)).
We therefore developed a new workflow to assemble ASVs from the long amplicons by splitting the reads into homologous domains, including the two ITS regions, 5.8S, the variable D1--3 regions of LSU [@michot1984], and the conserved LSU regions between the D regions, here referred to as LSU1--4 (Figure\ \@ref(fig:regions)a), followed by independently denoising reads from each domain, concatenating the denoised domains, and finally forming consensus sequences based on ITS2 identity.
We used this method, implemented in the new R packages `LSUx` (splitting reads into homologous regions; https://github.com/brendanf/LSUx) and `tzara` (reassembling regions and generating full-length consensus ASVs; https://github.com/brendanf/tzara) and detailed in Supplementary Methods \@ref(denoising-supplement), for all of the PacBio and Ion Torrent datasets.
Because the `LSUx` plus `tzara` method as currently implemented is not applicable to Illumina paired-end reads, the ASVs generated from the Illumina dataset according to the standard DADA2 workflow were used.
The ITS2 region was extracted from the ASVs using `LSUx` for comparison to the results from the other technologies.
To account for intra-species variation and the possibility of different denoising performance between the different sequencing strategies, the pooled ITS2-ASVs from all sequencing strategies were also clustered into operational taxonomic units (OTUs) at 97% similarity using VSEARCH v2.9.1 [@rognes2016].

@@

Full length long amplicon ASVs were aligned using DECIPHER [@wright2015] with up to 10\ iterations of alternating progressive alignment and conserved RNA secondary structure calculation, followed by 10\ refinement iterations.
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had introns after this position, as also observed in several fungal species by @holst-jensen1999.
An ML tree was produced using RAxML version 8.2.12 [@stamatakis2014] using the GTR+GAMMA model and rapid bootstrapping with the MRE_IGN stopping criterion.
Sequences confidently assigned outside kingdom Fungi by at least five of the primary taxonomic identification methods were used to root the tree, and sequences outside the clade defined by confidently identified Fungi were removed (see below and Supplementary Methods \@ref(rooting)).

@@

For full-length long-amplicon ASVs, the primary taxonomic assignments were refined with the new algorithm PHYLOTAX.
The PHYLOTAX algorithm resolves conflictsm one or more primary assignment methods (see Figure\ \@ref(fig:phylotax) and Supplementary Methods \@ref(phylotax)).
It is available in the new R package `phylotax` at https://github.com/brendanf/phylotax.

@@

Each of the distance matrices was used to calculate a Mantel correlogram with a 1\ m bin size for distances in the range of 0--12\ m, i.e., half the maximum separation present in the dataset.
Separate correlograms were drawn for samples taken during the same year and samples separated in time by one year, in order to assess the degree to which the soil community changes over the course of one year.
Additionally, empirical spatiotemporal distance-decay curves were generated by plotting mean community dissimilarity as a function of spatial distance and time lag, and fit to an exponential model of the form given by @legendre2012 using the `nls` function in R (Supplementary Methods \@ref(spatiotemporal-supplement)).
Spatiotemporal analyses were performed on the full recovered fungal community after removal of non-fungal sequences, and on the ECM community.

@@

```{r venn, fig.cap="(ref:venn)", fig.scap = "(ref:venn-s)", fig.show="hold", fig.height = 4, fig.align="center", results="asis"}
line = ggplot() +
  geom_vline(xintercept = 0) +
  theme_pubclean() +
  theme(axis.title = element_text(color = "white"),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()
        )
ggpubr::ggarrange(
  readd(vennplot_tech_ASV),
  line,
  readd(vennplot_tech_OTU),
  ncol = 3,
  widths = c(1,0.02,1),
  labels = c("ASVs", "", "OTUs"),
  label.x = 0.5,
  hjust = 0.5,
  font.label = list(size = 16)
)
grid::grid.legend(
  labels =  c("PacBio Short", "Illumina", "Ion Torrent"),
  nrow = 1,
  pch = 21,
  gp = grid::gpar(fill = c("grey85", "lightblue", "lightcoral")),
  vp = grid::viewport(y = unit(0.05, "npc"))
)

>>>>

Sampling was conducted at two sites, near the villages of Angaradebou (Ang: N 9.75456° E 2.14064°) and Gando (Gan: N 9.75678° E 2.31058°) approximately 30\ km apart in the _Forêt Classée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in West Sudanian savannah woodlands [@olson2001;@yorou2014] dominated by the ECM host tree _Isoberlinia doka_ (Fabaceae tr. Amherstieae).
At each site, 25\ soil samples were collected along a single 24\ m linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016, for a total of 67 samples.
For each sample, coarse organic debris was removed from the soil surface and a sample of approximately 5\ cm\ × 5\ cm\ × 5\ cm was extracted with an ethanol sterilized knife blade and homogenized by manually breaking up clumps and mixing inside a plastic zipper bag.
A subsample of approximately 250\ mg was stored in DNA-preservation buffer and returned to the laboratory for extraction (see Supplementary Methods \@ref(DNA-supplement) for preservation and extraction methods).

@@

We conclude that this poor performance was due to a combination of long amplicon length and low sequencing depth relative to community diversity, which lead to most biological variants being represented only by a cluster of reads differing by a small number of unique sequencing errors (for calculations see Supplementary Methods \@ref(long-numbers)).
We therefore developed a new workflow to assemble ASVs from the long amplicons by splitting the reads into homologous domains, including the two ITS regions, 5.8S, the variable D1--3 regions of LSU [@michot1984], and the conserved LSU regions between the D regions, here referred to as LSU1--4 (Figure\ \@ref(fig:regions)a).
We then independently denoised reads from each domain and concatenating the denoised domains for each read.
Finally denoised reads were clustered based on 100% ITS2 identity, and a consensus ASV was calculated for each cluster.
We used this method, implemented in the new R packages `LSUx` (splitting reads into homologous regions; https://github.com/brendanf/LSUx) and `tzara` (reassembling regions and generating full-length consensus ASVs; https://github.com/brendanf/tzara) and detailed in Supplementary Methods \@ref(denoising-supplement), for all of the PacBio and Ion Torrent datasets.
Because the `LSUx` plus `tzara` method as currently implemented is not applicable to Illumina paired-end reads, the ASVs generated from the Illumina dataset according to the standard DADA2 workflow were used.
The ITS2 region was extracted from the ASVs using `LSUx` for comparison to the results from the other technologies.
To account for intra-species variation and the possibility of different denoising performance between the different sequencing strategies, the pooled ITS2-ASVs from all sequencing strategies were also clustered into operational taxonomic units (OTUs) at 97% similarity using VSEARCH v2.9.1 [@rognes2016].

@@

Full length long amplicon ASVs were aligned using DECIPHER [@wright2015] with up to 10\ iterations of alternating progressive alignment and conserved RNA secondary structure calculation, followed by 10\ refinement iterations.
This alignment was truncated at a position after the D3 region corresponding to base 907 of the *Saccharomyces cerevisiae* S288C reference sequence for LSU, because several sequences had introns after this position, as also observed in several fungal species by @holst-jensen1999.
An ML tree was produced using RAxML version 8.2.12 [@stamatakis2014] using the GTR+GAMMA model and rapid bootstrapping with the MRE_IGN stopping criterion.
Sequences confidently (i.e., by at least five of the primary taxonomic identification methods) assigned outside kingdom Fungi  were used to root the tree, and sequences outside the clade defined by confidently identified Fungi were removed (see below and Supplementary Methods \@ref(rooting)).

@@

For full-length long-amplicon ASVs, the primary taxonomic assignments were refined with the new algorithm PHYLOTAX.
The PHYLOTAX algorithm resolves conflicts among one or more primary assignment methods (see Figure\ \@ref(fig:phylotax) and Supplementary Methods \@ref(phylotax)).
It is available in the new R package `phylotax` at https://github.com/brendanf/phylotax.

@@

Each of the distance matrices was used to calculate a Mantel correlogram with a 1\ m bin size for distances in the range of 0--12\ m, i.e., half the maximum separation present in the dataset.
Separate correlograms were drawn for samples taken during the same year and samples separated in time by one year, in order to assess the degree to which the soil community changes over the course of one year.
Additionally, empirical spatiotemporal distance-decay curves were generated by plotting mean community dissimilarity as a function of spatial distance and time lag, and fit to an exponential model of the form given by @legendre2012 using the `nls` function in R (Supplementary Methods \@ref(spatiotemporal-supplement)).
Spatiotemporal analyses were performed on the full recovered fungal community after removal of non-fungal sequences, and on the ECM community.
Sequences were assigned as ECM based on taxonomic assignments using the FUNGuild database [@nguyen2016funguild] via the R package `FUNGuildR` (https://github.com/brendanf/FUNGuildR).
All taxa which included "Ectomycorrhiza" in the guild assignment at any level of confidence were included.

@@

```{r venn, fig.cap="(ref:venn)", fig.scap = "(ref:venn-s)", fig.show="hold", fig.height = 4, fig.align="center", results="asis"}
line = ggplot() +
  geom_vline(xintercept = 0) +
  ggpubr::theme_pubclean() +
  theme(axis.title = element_text(color = "white"),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.background = element_blank(),
        panel.background = element_blank()
        )
ggpubr::ggarrange(
  readd(vennplot_tech_ASV),
  line,
  readd(vennplot_tech_OTU),
  ncol = 3,
  widths = c(1,0.02,1),
  labels = c("ASVs", "", "OTUs"),
  label.x = 0.5,
  hjust = 0.5,
  font.label = list(size = 16)
)
grid::grid.legend(
  labels =  c("PacBio Short", "Illumina", "Ion Torrent"),
  nrow = 1,
  pch = 21,
  gp = grid::gpar(fill = c("grey85", "lightblue", "lightcoral")),
  vp = grid::viewport(y = unit(0.05, "npc"))
)

### 2021-02-18 writing/transect_supplement.Rmd
Additional changes from Martin

The phylogenetic tree was rooted based on sequences which were confidently assigned to kingdom based on the primary taxonomic assignments.
Assignments based on Warcup were not used at this step because non-Fungi are not included in the dataset, and thus it cannot be used to directly assign kingdoms.
Assignments were considered reliable if at least five of the six remaining primary assignment methods (2 databases × 3 algorithms) resulted in the same kingdom-level assignment, with no conflicting assignments.
Two ASVs belonging to *Tulasnella*, which is known to have an accelerated rate of molecular evolution in the rDNA and often forms long, unstable branches in phylogenies [@moncalvo2006], were excluded at this step.
This resulted in `r nrow(readd(sure_kingdoms))` confident kingdom assignments, with `r dplyr::count(readd(sure_kingdoms), taxon) %>% glue::glue_data("{n} assigned to {taxon}") %>% glue::glue_collapse(", ", last = ", and ")`.
Inspection of the tree verified that these four kingdoms were reciprocally monophyletic, and the clade containing Alveolata and Viridiplantae was used as an outgroup to root the tree.
The kingdom Fungi was then identified as the minimal clade containing all ASVs which were confidently identified  as kingdom Fungi.
ASVs falling outside this clade, including the excluded *Tulasnella* sequences, were not included in downstream fungal community analysis.

>>>>

The phylogenetic tree was rooted based on sequences which were confidently assigned to kingdom based on the primary taxonomic assignments.
Assignments based on Warcup were not used at this step because non-Fungi are not included in the dataset, and thus it cannot be used to directly assign kingdoms.
Assignments were considered confident if at least five of the six remaining primary assignment methods (2 databases × 3 algorithms) resulted in the same kingdom-level assignment, with no conflicting assignments.
Two ASVs belonging to *Tulasnella*, which is known to have an accelerated rate of molecular evolution in the rDNA and often forms long, unstable branches in phylogenies [@moncalvo2006], were excluded at this step.
This resulted in `r nrow(readd(sure_kingdoms))` confident kingdom assignments, with `r dplyr::count(readd(sure_kingdoms), taxon) %>% glue::glue_data("{n} assigned to {taxon}") %>% glue::glue_collapse(", ", last = ", and ")`.
Inspection of the tree verified that these four kingdoms were reciprocally monophyletic, and the clade containing Alveolata and Viridiplantae was used as an outgroup to root the tree.
The kingdom Fungi was then identified as the minimal clade containing all ASVs which were confidently identified  as kingdom Fungi.
ASVs falling outside this clade, including the excluded *Tulasnella* sequences, were not included in downstream fungal community analysis.
